Dissertation Proposal

### 2023-11-09 proposal.tex
clean up high-level structures

@@

\section{Plan for \textbf{RQ1}}%
\label{s:plan-rq1}

@@

\subsection{Populations}%
\label{s:populations}

@@

\subsection{Find Misconceptions in Student-written Programs}%
\label{s:find-misconceptions-with-quizius}

@@

\subsection{Find Misconceptions with Expert-written Programs}%
\label{s:find-misconceptions-with-tutor}

@@

To be reasonably specific about misconceptions, I have created a
definitional interpreter for each misconception that I have found. I
plan to do the same things for any new misconceptions. I call these
interpreters \emph{misinterpreters}.

\section{Plan for \textbf{RQ2}}%
\label{s:plan-rq2}

@@

@@

\subsection{Tutorials and Language Levels}%
\label{s:tutorials}

There are many misconceptions to correct. To make the tutoring process
manageable, I have designed the tutor as a sequence of tutorials.
Inspired by prior research on language levels
(\Cref{s:pedagogic-techniques}), my sequence of tutorials is ordered,
and each tutorial covers one or more new language constructs
(\Cref{t:tutorials}).

Each tutorial might be delivered as multiple modules to avoid forcing
students to continuously work for more than 30 minutes.

@@

\subsection{Interpreting Tasks}%
\label{s:interpreting-tasks}

The tutor is mostly consisted of \emph{interpreting tasks}:
multiple-choice questions (MCQs) that asks students to interpret
program.

\paragraph{The MCQ Options} The options include (among other options
that I will explain shortly) the correct answer and wrong answers that
correspond to known misconceptions. Because students might conceive
wrong answers that I do not anticipate, the tutor has been providing
an ``Other'' option that, once chosen, allows students to enter
arbitrary answers.

@@

\subsection{Refutation Text and Other Explanations}%
\label{s:refutationt-text}

The tutor has been presenting a refutation text when a student chooses
a wrong answer that corresponds to known misconception(s). Let's
reconsider the program in \Cref{s:interpreting-tasks}. If a student
answers \lstinline|0 0|, which suggests they hold the
\miscon{DefByRef} misconception, the tutor responds
\begin{quote}
  \lstinline|y| was bound to \lstinline|12| (i.e., the value of
  \lstinline|x|) rather than to \lstinline|x|. So changing the value
  of \lstinline|y| does not change the value of \lstinline|x|.
\end{quote}
If a student gave a wrong answer that does not correspond to a
misconception (either by choosing one of the generated option or by
choosing the ``Other'' option), the tutor presents a generic
explanation. For example, if a student answers \lstinline|12 12|, the
tutor responds
\begin{quote}
  The first definition binds \lstinline|x| to \lstinline|12|. The
  second definition binds \lstinline|y| to the value of \lstinline|x|,
  which is \lstinline|12|. The \lstinline|set!| mutates the binding of
  \lstinline|y|, so \lstinline|y| is now bound to \lstinline|0|.
\end{quote}
If a student gave the correct answer, the tutor also presents the
generic explanation.

\subsection{Case Comparison}%
\label{s:case-comparison}

@@

\subsection{Notional Machine}%
\label{s:notional-machine}

@@

\subsection{Collating Problems for the Tutor}%
\label{s:tutor-refine-program-set}

@@

  \item I removed programs of a ``Lispy'' nature (See
        \Cref{s:reduce-syntactic-bias}).

@@

\subsection{Notional Machine}

@@

This approach is inconvenient both for the students and for me. For
students, they have to click, paste, and run, which is less convenient
than a single click. For me, because students run program in DrRacket
rather than ``within'' the tutor, I do not know whether they have run
the programs. This limits my ability to claim that the tutor is using
NM to help teach.

@@

The latest tutor provides a button on the top-right of each program.
If students click the button, they will open a new browser tab that
shows the NM populated with the program. I have deployed the latest
tutor to the latest population at Brown (i.e., University 1), but I
have not analyzed the data, which I plan to do in Spring 2024.

I have not changed the logging systems to log when students click the
buttons. I plan to make the change before Spring 2024.

Another issue with the current design is that students might run a
program before giving an answer. I don't know yet how often this
happens. But, given that students receive a penalty (\Cref{s:penalty})
for giving wrong answers, they might do so when the program is overly
complicated. This problem is more of an issue after the switch to a
web-based notional machine because now the ``cost'' of running a
program is cheaper. My plan is to change the tutor so that it provides
a program-specific link \emph{only} after each interpreting task.

\subsection{Reduce Syntactic Bias}

The processes described in
\Cref{s:find-misconceptions-with-quizius,s:find-misconceptions-with-tutor}
have been conducted with SMoL's Lispy syntax (\Cref{s:smol}). This can
potentially introduce biases to the found misconceptions. For example,
current results find that some students think the following program
produces \lstinline|'(4 5)| rather than \lstinline|'(1 2)|.
\begin{lstlisting}
  (filter (lambda (n) (> 3 n)) '(1 2 3 4 5))
\end{lstlisting}
This checks whether $\verb|3| > \verb|n|$, but those students
presumably vocalized it as ``greater than \lstinline|3|,
\lstinline|n|?''. Although I have tried to remove programs like this
from my instruments, there is still a risk of having biases in student
errors. This semester (Fall 2023) I am already deploying a newer SMoL
Tutor at Brown. This new tutor is able to display programs in
JavaScript and Python, in addition to the Lispy syntax. I plan to
compare in Spring 2024 the new tutor data with the current results,
i.e., a comparison between uni-syntax tutor and multi-syntax tutor.

\subsection{Use Case Comparisons More Consistently}

@@

\subsection{Options in Interpreting Tasks}

@@

\subsection{Other Changes in the Tutor}

By the end of Fall 2023, I also want to improve the SMoL Tutor at
least on the following aspects:
\begin{enumerate}
  \item The tutor currently does not ask students to explain when they
        gave a wrong answer in an interpreting question. This limits my
        ability to find new misconceptions and to confirm that the
        tasks do detect the anticipated misconceptions. I plan to
        change the tutor so that it asks for an explanation in each
        interpreting task.
        % With student explanations, I check whether
        % choosing the wrong answer indicates holding the corresponding
        % misconception(s). I plan to eyeball the explanations to find
        % new misconceptions. This piece of data also helps me confirm
        % that anticipated wrong answers represent the expected
        % misconceptions.
  \item As I said in \Cref{s:effect-tutor}, for many of the
        misconceptions, the tutor has too few questions. On the other
        hand, there are many questions that do not produce useful data
        because their wrong answers do not uniquely point to a
        misconception. I plan to adjust the collection of interpreting
        questions so that every misconception has at least two
        questions that aim to correct the misconception.
  \item The tutor is not randomizing the order of questions. This makes
        it difficult to compare questions for one misconception --
        earlier questions might be difficult simply because the
        programs are somehow more difficult. I plan to randomize the
        order of interpreting tasks within each tutorial module.
  \item After randomizing question orders, I would need to check
        carefully the dependencies between explanations and
        resolve these dependencies, if any.
  \item The tutor uses many variable names, some are separated by dashes
        (e.g., \lstinline|get-x|). Students might misread these
        variables. For example, some students might misread
        \lstinline|(get-x)| (a function named \lstinline|get-x| is
        called with no argument) as \lstinline|(get x)| (a function
        named \lstinline|get| is called with one argument
        \lstinline|x|). These misreadings add noise to the data. So I
        plan to use consistent and simple variable names in all
        programs.
\end{enumerate}

\subsection{Explore More Pedagogic Techniques}

\subsection{Expand Populations}

@@

\subsection{Further Plans}

@@

During Fall 2024, I plan to collect a new round of data from Brown, analyze
data that I collected in Spring 2024, and hopefully write a paper about the
updates.

>>>>

\clearpage

\tableofcontents

\clearpage

@@

\section{Plan for \textbf{RQ1}: Find Misconceptions}%
\label{s:plan-rq1}

@@

\subsection{In What Populations to Find?}%
\label{s:populations}

@@

\subsection{Find Misconceptions by Analyzing Student-written Programs}%
\label{s:find-misconceptions-with-quizius}

@@

\subsection{Find Misconceptions by Testing Students with Expert-written Programs}%
\label{s:find-misconceptions-with-tutor}

@@

To be reasonably specific about misconceptions, I have created a
definitional interpreter for each misconception that I have found. I
plan to do the same things for any new misconceptions. I call these
interpreters \emph{misinterpreters}. Having definitional interpreters
makes it convenient (and less error-prone) to associate wrong answers 
with misconceptions (\Cref{s:misconceptions-tutor}).

\section{Plan for \textbf{RQ2}: Correct Misconceptions}%
\label{s:plan-rq2}

@@

To answer this RQ, I plan to create a tutor (\Cref{s:design-tutor}) and
argue that the tutor is effective (\Cref{s:evaluation}).

\subsection{Design a Tutor for Correcting Misconceptions}%
\label{s:design-tutor}

@@

\subsubsection{Tutorials and Language Levels}%
\label{s:tutorials}

There are many misconceptions to correct. To make the tutoring process
manageable, I have designed the tutor as a sequence of tutorials.
Inspired by prior research on language levels
(\Cref{s:pedagogic-techniques}), my sequence of tutorials is ordered,
and each tutorial covers one or more new language constructs
(\Cref{t:tutorials}). Each tutorial might be delivered as multiple modules to avoid forcing
students to continuously work for more than 30 minutes.

@@

\subsubsection{Interpreting Tasks}%
\label{s:interpreting-tasks}

The tutor is mostly consisted of \emph{interpreting tasks}:
multiple-choice questions (MCQs) that asks students to interpret
program. The options include (among other options
that I will explain shortly) the correct answer and wrong answers that
correspond to known misconceptions. Because students might conceive
wrong answers that I do not anticipate, the tutor has been providing
an ``Other'' option that, once chosen, allows students to enter
arbitrary answers.

@@

\subsubsection{Design with Pedagogic Techniques Known to Work}

\paragraph{Refutation Text and Other Explanations}%
\label{s:refutationt-text}
The tutor has been presenting a refutation text when a student chooses
a wrong answer that corresponds to known misconception(s). Let's
reconsider the program in \Cref{s:interpreting-tasks}. If a student
answers \lstinline|0 0|, which suggests they hold the
\miscon{DefByRef} misconception, the tutor responds
\begin{quote}
  \lstinline|y| was bound to \lstinline|12| (i.e., the value of
  \lstinline|x|) rather than to \lstinline|x|. So changing the value
  of \lstinline|y| does not change the value of \lstinline|x|.
\end{quote}
If a student gave a wrong answer that does not correspond to a
misconception (either by choosing one of the generated option or by
choosing the ``Other'' option), the tutor presents a generic
explanation. For example, if a student answers \lstinline|12 12|, the
tutor responds
\begin{quote}
  The first definition binds \lstinline|x| to \lstinline|12|. The
  second definition binds \lstinline|y| to the value of \lstinline|x|,
  which is \lstinline|12|. The \lstinline|set!| mutates the binding of
  \lstinline|y|, so \lstinline|y| is now bound to \lstinline|0|.
\end{quote}
If a student gave the correct answer, the tutor also presents the
generic explanation.

\paragraph{Case Comparison}%
\label{s:case-comparison}

@@

\paragraph{Notional Machine}%
\label{s:notional-machine}

@@

\subsection{Further Collating Problems, for the Tutor}%
\label{s:tutor-refine-program-set}

@@

  \item I removed programs of a ``Lispy'' nature.

@@

\subsection{Keep Exploring More Pedagogic Techniques}

\subsection{Implemented But Not Evaluated Improvements in the Tutor}

\paragraph{Ease Access to the Notional Machine}

@@

This approach is inconvenient both for the students and for me. For
students, they have to click, paste, and run, which is less convenient
than a single click.

@@

The latest tutor provides a button on the top-right of each program.
If students click the button, they will open a new browser tab that
shows the NM populated with the program. I have deployed the latest
tutor to the latest population at Brown (i.e., University 1), but I
have not analyzed the data, so the new tutor is not presented in
the preliminary results.

\paragraph{Reduce Syntactic Bias with Multiple Syntax}

The processes described in
\Cref{s:find-misconceptions-with-quizius,s:find-misconceptions-with-tutor}
have been conducted with SMoL's Lispy syntax (\Cref{s:smol}). This can
potentially introduce biases to the found misconceptions. For example,
when using SMoL Quizzes, I found that some students think the following program
produces \lstinline|'(4 5)| rather than \lstinline|'(1 2)|.
\begin{lstlisting}
  (filter (lambda (n) (> 3 n)) '(1 2 3 4 5))
\end{lstlisting}
This checks whether $\verb|3| > \verb|n|$, but those students
presumably vocalized it as ``greater than \lstinline|3|,
\lstinline|n|?''. Although I have tried to remove programs like this
from the tutor, there is still a risk of having biases in student
errors. This semester (Fall 2023) I am already deploying a newer SMoL
Tutor at Brown. This new tutor is able to display programs in
JavaScript and Python, in addition to the Lispy syntax. I plan to
compare in Spring 2024 the new tutor data with the current results,
i.e., a comparison between uni-syntax tutor and multi-syntax tutor.

\subsection{The Rest of Fall 2023: Improve the Tutor}

\paragraph{Log Access to the Notional Machine}

I have not changed the logging systems to log when students click the
buttons to access the NM. I plan to make the change before Spring 2024.

\paragraph{Limit Access to the Notional Machine}

Another issue with the current design is that students might run a
program before giving an answer. I don't know how often this
happens. But, given that students receive a penalty (\Cref{s:penalty})
for giving wrong answers, they might do so when the program is overly
complicated. This problem is more of an issue after the switch to a
web-based notional machine because now the ``cost'' of running a
program is cheaper. My plan is to change the tutor so that it provides
a program-specific link \emph{only} after each interpreting task.

\paragraph{Use Case Comparisons More Consistently}

@@

\paragraph{Use More Sensible Options in Interpreting Tasks}

@@

\paragraph{Ask For Explanation on Choices}

The tutor currently does not ask students to explain when they
        gave a wrong answer in an interpreting question. This limits my
        ability to find new misconceptions and to confirm that the
        tasks do detect the anticipated misconceptions. I plan to
        change the tutor so that it asks for an explanation in each
        interpreting task.

\paragraph{Clean Up the Program Set}
        % With student explanations, I check whether
        % choosing the wrong answer indicates holding the corresponding
        % misconception(s). I plan to eyeball the explanations to find
        % new misconceptions. This piece of data also helps me confirm
        % that anticipated wrong answers represent the expected
        % misconceptions.
As I said in \Cref{s:effect-tutor}, for many of the
        misconceptions, the tutor has too few questions. On the other
        hand, there are many questions that do not produce useful data
        because their wrong answers do not uniquely point to a
        misconception. I plan to adjust the collection of interpreting
        questions so that every misconception has at least two
        questions that aim to correct the misconception.

\paragraph{Randomize the Order of Interpreting Tasks}

The tutor is not randomizing the order of questions. This makes
        it difficult to compare questions for one misconception --
        earlier questions might be difficult simply because the
        programs are somehow more difficult. I plan to randomize the
        order of interpreting tasks within each tutorial module.
After randomizing question orders, I would need to check
        carefully the dependencies between explanations and
        resolve these dependencies, if any.

\paragraph{Clean Up Variable Names}

The tutor uses many variable names, some are separated by dashes
        (e.g., \lstinline|get-x|). Students might misread these
        variables. For example, some students might misread
        \lstinline|(get-x)| (a function named \lstinline|get-x| is
        called with no argument) as \lstinline|(get x)| (a function
        named \lstinline|get| is called with one argument
        \lstinline|x|). These misreadings add noise to the data. So I
        plan to use consistent and simple variable names in all
        programs.

\subsection{Spring 2024: Collect More Data}

@@

@@

\subsection{Summer 2024: Data Analysis and Write a Paper}

During this time, I plan to analyze the data I collected in Fall 2023 
and Spring 2024 and start working on paper about the changes in the tutor.

\subsection{Fall 2024: Collect More Data and Write a Paper}

During Fall 2024, I plan to collect a new round of data from Brown, analyze
data that I collected in Spring 2024, and hopefully finish writing the new paper.

\subsection{Spring 2025: Defense}
